```{r}
#| label: import-packages
#| echo: false
#| include: false
library(tidyverse)
library(survey)


```

```{r}
#| label: import-arenas
#| echo: false
#| include: false
arenas = readxl::read_excel("~/sta322_methods3/data/arenas.xlsx")
arenas = drop_na(arenas)
```
## 1. 

### a. 
For our margin of error, we are given the formula, 

$$
MOE \approx 2 \sqrt{\frac{(1 - \frac{n}{N})S^2}{n}} . 
$$

Then, we are told to estimate $S$ with $S_{guess} = 5000$ and we want $n$ to produce a MOE around 1000, i.e. find an $n$ such that

$$
1000 = 2 \sqrt{\frac{(1 - \frac{n}{N})5000^2}{n}} 
$$

$$
500 =  \sqrt{\frac{(1 - \frac{n}{N})5000^2}{n}} \\
$$

$$
250,000 = \frac{(1 - \frac{n}{N})5000^2}{n} \\
$$

$$
250,000 \times n = 5000^2 - 5000^2n \frac{1}{N} \\
$$

$$
n(250,000+\frac{5000^2}{N}) = 5000^2 \\
$$

$$
n = \frac{5000^2}{250,000+\frac{5000^2}{N}}, \quad \text{given} ~~ N=383
$$

```{r}
#| label: 1a
#| echo: false

n_moe = (5000^2)/(250000+((5000^2)/383))
paste0("Calculated n: ", n_moe)
```
$$
n = 79.296... \approx 80
$$


Check:
```{r}
n = 80
N = 383
moe_n80 = 2 * sqrt((1 - n / N) * 5000^2 / n)
```

```{r}
#| echo: FALSE

paste0("The MOE we get from sample size of 80 is ", round(moe_n80, digits=1), ", roughly rounding to 1000")
```

### b.

```{r}
#| label: 1b
set.seed(0)
sampled_rows = sample(1:nrow(arenas), 80)
arena_srs = arenas[sampled_rows, ]
```

### c. 

```{r}
#| label: 1c

n=80
N=383

arena_srs$wt = n/N #equal weight for each observation in SRS
arena_srs$fpc = 383
srs_des = svydesign(~1, weights=~wt, fpc=~fpc, data=arena_srs)

svymean(~Capacity, srs_des, na.rm=TRUE)
confint(svymean(~Capacity, srs_des,  na.rm = TRUE), level=.95)

```
We have a decently large sample size of 80, so central limit theory should be applicable. As a safety measure, we can check to see if `Capacity` behaves weirdly i.e. would need a larger sample size for the sample mean distribution to resemble a bell curve.
```{r}
#| echo: FALSE
hist(arena_srs$Capacity)
```
While there is some skew, it is not extreme. Therefore, we can use our confidence interval to say that:

"With a point estimate of 7182.1 as our average capacity, we are 95% sure that $$\bar{Y} \in (6193.06, 8006.51)$$"

### d. 

```{r}
#| label: 1d

arena_srs$Less = as.numeric(arena_srs$Capacity < 9314)
srs_des = svydesign(~1, weights=~wt, fpc=~fpc, data=arena_srs)

arena_less = svymean(~Less,
  design = srs_des,
  na.rm = TRUE)
```



The estimated percentage of arenas with capacity below 9314 is obtained by taking the proportion of arenas in the sample that meet that condition. We did this by making the indicator variable `Less`, which is 1 for a capacity below 9314 and 0 otherwise. Then, by taking the SRS mean, we effectively get a sample proportion, our estimator.  

Central limit theorem is applicable here, but we may need a larger sample size if we have a heavy skew in the sample proportion distribution. This would mean that if $p$ is close to either 0 or 1, it might be risky to use a confidence interval using SE and critical value. But note that this likely isn't the case, as our sample proportion is around 0.76. 

So, we can declare 95% confidence interval:

```{r}
arena_less
confint(arena_less, level = 0.95)
```

> (0.67, 0.84) with 0.75 as our estimate.


## 2.

### a.

```{r}
#| label: 2a

arenas = arenas |>
  mutate(
    strat = case_when(
      Opened < 1960 ~ "Classics",
      Opened < 1980 ~ "Older",
      Opened < 2000 ~ "Middle",
      TRUE          ~ "New"
    ),
    strat = factor(strat, levels = c("Classics", "Older", "Middle", "New"))
  )

table(arenas$strat)
```

With allocation, the stratum sample size is $n_h = n \times W_h$, where $W_h = N_h / N$. These are not integers, so we
round such that `n_h` is rounded down, except for the strata with the largest remainder. That strata gets the "leftover" in a sense. This ensures that
$\sum_h n_h = 80$ while staying as close to the proportions as possible. 

```{r}
#| label: 2a-allocation

#we need a function to deal with an uneven split of sample, function logic came from Claude
largest_remainder = function(exact, total) {
  fl = floor(exact)
  remain = total - sum(fl)
  if (remain > 0) {
    ord = order(exact - fl, decreasing = TRUE)
    fl[ord[seq_len(remain)]] = fl[ord[seq_len(remain)]] + 1
  }
  fl
}

Nh = table(arenas$strat)
Wh = Nh/N

exact_nh = n * Wh
nh = largest_remainder(as.numeric(exact_nh), n)
names(nh) = names(Nh)

allocation_table = tibble(
  stratum  = names(Nh),
  Nh       = as.numeric(Nh),
  Wh       = round(as.numeric(Wh), 4),
  exact_nh = round(as.numeric(exact_nh), 2),
  nh       = nh
)

allocation_table
sum(nh)  #must be 80
```

Now, we can draw a SRS stratified sample.
`slice_sample()` samples without replacement by default, and grouping by `strat` first means
the draws happen independently within each stratum.

```{r}
#| label: 2a-sample

set.seed(322)

arena_str = arenas |>
  group_by(strat) |>
  group_modify(~ slice_sample(.x, n = nh[[as.character(.y$strat)]])) |>
  ungroup()

count(arena_str, strat)
```

### b.

Each arena in stratum $h$ is selected with equal proability $n_h / N_h$, so each arena has a  sampling
weight of $N_h / n_h$. 
```{r}
#| label: 2b

arena_str = arena_str |>
  mutate(
    strwt  = as.numeric(Nh[as.character(strat)]) / nh[as.character(strat)],
    strfpc = as.numeric(Nh[as.character(strat)])
  )

str_des = svydesign(~1, strata  = ~strat, weights = ~strwt, fpc = ~strfpc, data = arena_str)

str_mean = svymean(~Capacity, str_des, na.rm = TRUE)
str_mean
confint(str_mean, level = 0.95)
```

With a total sample size of $n = 80$ and no extreme skew in capacity, the
central limit theorem is reasonable here, so the interval (6992.1, 8843.6) works as a 95% confidence interval for $\bar{Y}$.

### c.


```{r}
#| label: 2c

arena_str = arena_str |>
  mutate(Less = as.numeric(Capacity < 9314))

str_des = svydesign(
  ~1,
  strata  = ~strat,
  weights = ~strwt,
  fpc     = ~strfpc,
  data    = arena_str
)

str_prop = svymean(~Less, str_des, na.rm = TRUE)
str_prop
confint(str_prop, level = 0.95)
```

The estimated proportion is not close to either 0 or 1, so the central limit theorem is suitable. 
Therefore, we can use 95% confidence interval (0.59, 0.78)

## 3.

Optimal allocation  follows the formula:

$$
n_h = n \cdot \frac{N_h S_h}{\sum_{k=1}^{H} N_k S_k},
$$

such that strata that are both large and varying get more of the sample.
We substitute the sample within strata with standard deviations $s_h$ from problem 2
for the unknown $S_h$.

```{r}
#| label: 3-optimal

sh_tab = arena_str |>
  group_by(strat) |>
  summarise(
    nh_2 = n(),
    sh = sd(Capacity),
  ) |>
  ungroup() |>
  mutate(Nh = as.numeric(Nh[as.character(strat)]),
         Nh_sh = Nh*sh)

exact_opt = n*sh_tab$Nh_sh/sum(sh_tab$Nh_sh)
nh_opt = largest_remainder(exact_opt, n) #reuse rounding system from earlier

optimal_table = tibble(
  stratum   = as.character(sh_tab$strat),
  Nh        = sh_tab$Nh,
  s_h       = round(sh_tab$sh, 1),
  Nh_sh     = round(sh_tab$Nh_sh, 1),
  exact_nh  = round(exact_opt, 2),
  nh_optimal = nh_opt
)

optimal_table
sum(nh_opt)  # again, has to be 80
```

The same largest-remainder rounding guarantees the integer sizes add to 80, which checks out!

## 4.

Budget \$100000 with fixed costs \$10,000 leaves \$90000 of
variable budget. The population variances are assumed to be similar, so they'll cancel out of the allocation formula.

### a.

Every interview costs the same $\$75$, so no preference between
groups. With $S_1 = S_2$, optimal allocation reduces to proportional allocation.

Total sample size:

$$
n = \frac{90000}{75} = 1200.
$$

Splitting proportionally:

$$
n_1 = 0.95 \times 1200 = 1140, \qquad n_2 = 0.05 \times 1200 = 60.
$$

So we conclude **1140 internet households and 60 non-internet households**, using the full
$\$90000$.

### b.

Now the costs differ: $c_1 = \$15$ and $c_2 = \$100$. Optimal allocation under
unequal costs is

$$
n_h \propto \frac{W_h S_h}{\sqrt{c_h}},
$$

and with $S_1 = S_2$ this becomes $n_h ~\text{proportional to}~ W_h / \sqrt{c_h}$:

$$
\frac{n_1}{n_2} = \frac{0.95/\sqrt{15}}{0.05/\sqrt{100}}
= \frac{0.24527}{0.005} = 49.05.
$$

When we impose the budget constraint $15 n_1 + 100 n_2 = 90000$ with $n_1 = 49.05\,n_2$:

$$
n_2 \left( 49.05 \times 15 + 100 \right) = 90000
$$
$$
n_2 \times 835.8 = 90000 \quad \Longrightarrow \quad n_2 = 107.7
$$
$$
n_1 = 49.05 \times 107.7 = 5282.1
$$

```{r}

ratio = (0.95 / sqrt(15)) / (0.05 / sqrt(100))
n2 = 90000 / (ratio * 15 + 100)
n1 = ratio * n2
c(n1 = n1, n2 = n2)

15 * 5282 + 100 * 107
```

Rounding down, we get about 5282 internet households online and
107 non-internet households in person, costing $\$89930$. Cheap online
interviews buy a far larger total sample, which is why
the mixed-mode design does a better job with precision.

## 5.

The file is a stratified sample of trucks. `STRATUM` identifies each record's
stratum and `TABTRUCKS` is the tabulation weight. Since the weight is already the inverse selection
probability, it goes straight into `weights=`.

The population is roughly 85 million trucks against a sample
of a few thousand, so every $n_h/N_h$ is very close to zero, so it makes sense to treat our sampling as if with replacement.
```{r}
#| label: 5-setup
#| message: false

library(tidyverse)
library(survey)

trucks = read.csv("~/sta322_methods3/data/trucks.csv")

trucks = trucks |>
  mutate(
    one = 1,
    TruckType = factor(
      TRUCKTYPE,
      levels = 1:5,
      labels = c(
        "Pickups",
        "Minivans, other light vans, and SUVs",
        "Light single-unit trucks < 26,000 lbs",
        "Heavy single-unit trucks >= 26,000 lbs",
        "Truck-tractors"
      )
    )
  )

truck_des = svydesign(
  ids     = ~1,
  strata  = ~STRATUM,
  weights = ~TABTRUCKS,
  data    = trucks
)
```

### a. 

```{r}
#| label: 5a

N_hat <- svytotal(~one, truck_des)
N_hat
confint(N_hat, level = 0.95)
```

This returns roughly $8.52 \times 10^7$ trucks, matching the direct sum
`sum(trucks$TABTRUCKS)`.

The standard deviation is also roughly zero because, in a stratified design, the
weight attached to an observation in $h$th stratum is $w_{hi} = N_h / n_h$, so summing the
weights over the whole sample gives

$$
\hat{N} = \sum_{h=1}^{255} \sum_{i \in \mathcal{S}_h} \frac{N_h}{n_h}
= \sum_{h=1}^{255} n_h \cdot \frac{N_h}{n_h}
= \sum_{h=1}^{255} N_h = N .
$$

Therefore, every possible sample gets us the same $N$ estimate. The stratum
sample sizes are already set in the survey design, and the response variable is
the constant 1, so there is no randomness from our responses in our sample.
As a result, $\hat{N}$ gets the actual population quantity i.e. $s_h^2 = 0$, so every
term of $\sum_h N_h^2 (1 - n_h/N_h) s_h^2 / n_h$ is just 0.


### b.

```{r}
#| label: 5b

total_miles = svytotal(~MILES_ANNL, truck_des, na.rm = TRUE)
total_miles
confint(total_miles, level = 0.95)
```

The estimated total miles driven by all trucks in the population is about
$1.115 \times 10^{12}$, with a 95% confidence interval of roughly
$1.102 \times 10^{12}$ to $1.127 \times 10^{12}$ miles.

The sample is large, so Central Limit Theorem / using the confidence interval is appropiate.

### c. 
`svyby` applies a survey estimator separately within each level of a grouping
variable, carrying the design information into each subgroup. This is the right
tool here — subsetting the data frame by hand and re-running `svytotal` would
discard the stratum structure.

There is a convenient feature of this particular design: `trucktype` is one of the
two stratification variables (the 255 strata are 51 states $\times$ 5 truck
classes). Each truck class is therefore an exact union of 51 whole strata, not a
domain that cuts across them. The sample size within each class is fixed by the
design rather than random, so these are ordinary stratified totals and no
domain-estimation correction is needed.

```{r}
#| label: 5c

miles_by_type = svyby(
  ~MILES_ANNL, by = ~TruckType, design = truck_des, FUN     = svytotal, na.rm   = TRUE)

miles_by_type
confint(miles_by_type, level = 0.95)
```


## 6.

### a.

The Horvitz–Thompson estimator of a population total weights each sampled unit by
the inverse of its selection probability:

$$
\hat{t}_y = \sum_{h=1}^{5} \sum_{i \in \mathcal{S}_h} \frac{y_{hi}}{\pi_{hi}},
\qquad \pi_{hi} = \frac{n_h}{N_h}.
$$

For $h = 1, 2, 3$ we have $N_h = 1$. A stratum of size 1 forces $n_h = 1$: the
single unit must be in the sample. So $\pi_{hi} = 1/1 = 1$ and each of those
units contributes its own value, $y_h / 1 = y_h$.

For $h = 4, 5$ the selection probability is $n_h / N_h$, so the weight is
$N_h / n_h$ and the stratum contributes

$$
\sum_{i \in \mathcal{S}_h} \frac{N_h}{n_h} y_{hi}
= \frac{N_h}{n_h} \sum_{i \in \mathcal{S}_h} y_{hi}
= N_h \bar{y}_h .
$$

Summing the five strata:

$$
\hat{t}_y = y_1 + y_2 + y_3 + N_4 \bar{y}_4 + N_5 \bar{y}_5 .
$$

### b.

The first friend is correct.

The reason being that strata are sampled independently, so the variance of $\hat{t}_y$ is the sum of the five stratum-specific variances:

$$
\widehat{V}(\hat{t}_y) = \sum_{h=1}^{5} N_h^2 \left( 1 - \frac{n_h}{N_h} \right) \frac{s_h^2}{n_h}.
$$

For $h = 1, 2, 3$ the whole strata is sampled such that $n_h = N_h = 1$, so the expression $\left( 1 - \frac{n_h}{N_h} \right)$ is $1 - 1/1 = 0$ and the whole term is evaluated as 0. In other words, there is no variability in $y_1$, $y_2$, $y_3$.Dropping them is not throwing
away information-- the sample already matches that stratum within the population.

That leaves

$$
\widehat{V}(\hat{t}_y) = N_4^2 \left( 1 - \frac{n_4}{N_4} \right) \frac{s_4^2}{n_4}
+ N_5^2 \left( 1 - \frac{n_5}{N_5} \right) \frac{s_5^2}{n_5},
$$

which is the first friend's formula.


Then, we know the second friend is wrong. This is because pooling all $n$ observations
into one sample computes $s^2$ across the whole sample, meaning it doesn't make sense to have
differences in the $y_{hi}$ for each stratum. Then, the formula
$(1 - n/N) s^2 / n$ is the variance of an SRS *mean*, not for the estimator of the total $\hat{t}_y$, which then applies the wrong weights to strata 1–3 by treating the sampling as a random process, when the sampling of the one individual in each of the strata is already for certain. So, the second student gets an overestimate of the variance.
